Spatiotemporal Optimization of Cosmological Particel-Mesh Simulation

### 2022-12-28 docs/papers/sto/sto.tex
Draft: add obj gradient equation

\usepackage{microtype}
\usepackage{newpxtext, eulerpx}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{booktabs}

@@

Our observable is a snapshot $\hat \vz = (\hat\vx, \hat\vp)^\intercal$ at any
given $a$, which in general can be given by the interpolation of two snapshots
$\vz_{k-1}$ and $\vz_k$.
Following the discussion in \citet{Li2022a}, the observe operator in the adjoint
equation time stepping from $a_i$ to $a_{i-1}$ now reads
\begin{align}
  O_i^{i-1}: \qquad \vxi_{i-1} :\:\!\!\!&= \vxi_{i-1}
  + \frac{\p\cJ}{\p\hat\vx}\frac{\p\hat\vx}{\p\vx_{i-1}} \,, \nonumber\\
                    \vpi_{i-1} :\:\!\!\!&= \vpi_{i-1}
  + \frac{\p\cJ}{\p\hat\vp}\frac{\p\hat\vp}{\p\vp_{i-1}} \,,
\end{align}
where $(\vxi, \vpi)$ are the adjoint variables of $(\vx, \vp)$, and
${\p\hat\vz}/{\p\vz_i}$ is zero except for the relevant $\vz_{k-1}$ and $\vz_k$.
Similarly we have in the initial conditions of the adjoint equations
\begin{align}
  \vxi_n &= \frac{\p\cJ}{\p\hat\vx}\frac{\p\hat\vx}{\p\vx_n} \,, \nonumber\\
  \vpi_n &= \frac{\p\cJ}{\p\hat\vp}\frac{\p\hat\vp}{\p\vp_n} \,.
\end{align}

@@

@@

@@

\pmwd\ is open-source on
\href{https://github.com/eelregit/pmwd}{GitHub}, including the
\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/sto}{source
and scripts of this paper}.

>>>>

\usepackage{microtype}
\usepackage{newpxtext, eulerpx}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{nicefrac}

@@

Our observable is a snapshot $\hat \vz = (\hat\vx, \hat\vp)^\intercal$ at any
given $a$, which in general can be given by the interpolation of two snapshots
$\vz_{k-1}$ and $\vz_k$.
Following the discussion in \citet{Li2022a}, the observe operator in the adjoint
equation time stepping from $a_i$ to $a_{i-1}$ now reads
\begin{align}
  O_i^{i-1}: \qquad \vxi_{i-1} :\:\!\!\!&= \vxi_{i-1}
  + \frac{\p\cJ}{\p\hat\vx} \cdot \frac{\p\hat\vx}{\p\vx_{i-1}} \,, \nonumber\\
                    \vpi_{i-1} :\:\!\!\!&= \vpi_{i-1}
  + \frac{\p\cJ}{\p\hat\vp} \cdot \frac{\p\hat\vp}{\p\vp_{i-1}} \,,
\end{align}
where $\vlambda = (\vxi, \vpi)$ are the adjoint variables of $\vz = (\vx,
\vp)^\intercal$, and ${\p\hat\vz}/{\p\vz_i}$ is zero except for the relevant
$\vz_{k-1}$ and $\vz_k$.
Similarly we have in the initial conditions of the adjoint equations
\begin{align}
  \vxi_n &= \frac{\p\cJ}{\p\hat\vx} \cdot \frac{\p\hat\vx}{\p\vx_n} \,, \nonumber\\
  \vpi_n &= \frac{\p\cJ}{\p\hat\vp} \cdot \frac{\p\hat\vp}{\p\vp_n} \,.
\end{align}
In the reverse time evolution of the adjoint equations, these adjoint varibales
remain zero until the time stepping from $a_{k+1}$ to $a_k$.

@@

As derived in \citet{Li2022a}, the gradient of the objective function
$\cJ(\hat\vz(\vz_{k-1}, \vz_k, \vtheta), \theta)$ is accumulated through
\begin{multline}
  \frac{\d\cJ}{\d\vtheta} = \frac{\p\cJ}{\p\vtheta}
  %
  + \frac{\p\cJ}{\p\hat\vx} \cdot \frac{\p\hat\vx}{\p\vtheta}
  + \frac{\p\cJ}{\p\hat\vp} \cdot \frac{\p\hat\vp}{\p\vtheta}
  %
  + \vxi_0 \cdot \frac{\p\vx_0}{\p\vtheta}
  + \vpi_0 \cdot \frac{\p\vp_0}{\p\vtheta} \\
  %
  - \sum_{i=1}^n \Bigl[
    \bigl( \vpi_i \cdot\va_i \frac\p{\p\vtheta} + \vzeta_i
      \bigr) K(t_i, t_i, t_{i-\half}) \\
    + \vxi_i \cdot \vp_{i-\half}
      \frac{\p D(t_{i-\half}, t_i, t_{i-1})}{\p\vtheta} \\
    + \bigl( \vpi_{i-1} \cdot \va_{i-1} \frac\p{\p\vtheta} + \vzeta_{i-1}
      \bigr) K(t_{i-1}, t_{i-\half}, t_{i-1})
    \Bigr],
\label{KDFKO_grad}
\end{multline}
where $\vtheta$ could include cosmological parameters, initial conditions, as
well as STO parameters in this work.

@@

We use a B-spline to model the function $a(t)$ that maps uniformly sampled $t$
values to optimally distributed $a$ time steps.
The spline is parameterized with a few control points that are updated during
training.

@@

\pmwd\ \citep{Li2022b} is open-source on
\href{https://github.com/eelregit/pmwd}{GitHub}, including the
\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/sto}{source
and scripts of this paper}.
